Stage-Splitting a Modal Language

### 2014-10-14 paper/example.tex
removed staging annotations from the informal example

>>>>

\subsection{Splitting of Quickselect}

Here, we will manually split apart the \texttt{qSelect} function in order to give some intuition about how the operation works.

We start with the basic \texttt{qSelect}:

\begin{lstlisting} 
1`datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ... `

fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in`2`
      case compare k n of
        LT => `qSelect 1`left `2`k
      | EQ => h
      | GT => `qSelect 1`right `2`(k-n-1)`
\end{lstlisting}

Again our goal is to split this into \texttt{qSelect1} and \texttt{qSelect2} such that \texttt{qStaged l i = qSelect2 (qSelect1 l) i}.  
The first step in this process is to assume this conclusion at the recursive sites:

\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in`2`
      case compare k n of
        LT => qSelect2 (`1`qSelect1 left`2`) k
      | EQ => h
      | GT => qSelect2 (`1`qSelect1 right`2`) (k-n-1)`
\end{lstlisting}

We then pull the stage \bbone\ portion of each branch of the comparison up into a single tuple.

\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      let p = `1`(h,qSelect1 left,qSelect1 right) in`2`
      case compare k n of
        LT => qSelect2 (pi2 p) k
      | EQ => pi1 p
      | GT => qSelect2 (pi3 p) (k-n-1)`
\end{lstlisting}

We continue to pull up the precomputation and include more in it:

\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let p =
        let (left,right,n) = partition h t in
        `1`(n,h,qSelect1 left,qSelect1 right)
      in`2`
      case compare k (pi1 p) of
        LT => qSelect2 (pi3 p) k
      | EQ => pi2 p
      | GT => qSelect2 (pi4 p) (k-n-1)`
\end{lstlisting}

This completes the splitting of the cons branch.  
The empty branch can also be split, though in a trivial way:

\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => let p = () in `2`0`1`
  | Cons (h,t) => 
      let p =
        let (left,right,n) = partition h t in
        `1`(n,h,qSelect1 left,qSelect1 right)
      in`2`
      case compare k (pi1 p) of
        LT => qSelect2 (pi3 p) k
      | EQ => pi2 p
      | GT => qSelect2 (pi4 p) (k-n-1)`
\end{lstlisting}

To continue splitting the case statement, we need to us a sum type in the precomputation
to remind the code what branch its in.

\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  let p = 
    case l of
      Empty => Top ()
    | Cons (h,t) =>
        let (left,right,n) = partition h t in`1`
        Bot (n,h,qSelect1 left,qSelect1 right)
  in`2`
  case p of 
    Top () => 0
  | Bot p => 
      case compare k (pi1 p) of
        LT => qSelect2 (pi3 p) k
      | EQ => pi2 p
      | GT => qSelect2 (pi4 p) (k-n-1)`
\end{lstlisting}

We finish where we start, splitting \texttt{qSelect} into
\texttt{qSelect1} and \texttt{qSelect2}.  
Omitting a few type annotations, that is:

\begin{lstlisting} 
1`fun qSelect1 (l : list) = 
    case l of
      Empty => Top ()
    | Cons (h,t) =>
        let (left,right,n) = partition h t in
        Bot (n,h,qSelect1 left,qSelect1 right)
`2`
fun qSelect2 p (k : int) : int = 
  case p of 
    Top () => 0
  | Bot p => 
      case compare k (pi1 p) of
        LT => qSelect2 (pi3 p) k
      | EQ => pi2 p
      | GT => qSelect2 (pi4 p) (k-n-1)`
\end{lstlisting}
